---
title: "Problem Set 4"
subtitle: "Due date: 14 October"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
---

```{r}
library(tidyverse)
```

Please upload your completed assignment to the ELMs course site (under the assignments menu) and to your class Github repository. You need to upload both your Quarto document (with R code, as needed) and its rendered output (either as a PDF or HTML file).

**Total points: 30**

## Question 1

*Points: 5*

Suppose I am playing poker with a regular deck of 52 cards. I have three cards in my hand, one of which is a king. What is the probability that the next two cards I draw will both be kings?

```{r}
prob <- (3/49) * (2/48)
prob
```

Answer: The probability would be 0.00255102

## Question 2

*Points: 5*

A survey of 1,500 randomly selected individuals reveals that 46% of them have a college degree and that 31% of them have a gun in their house. If you were to assume that college graduation and gun ownership are independent of one another, what is the probability that an individual is both a college graduate and does not have a gun in their house?

```{r}
prob_of_nogun_and_college <- (0.46) * (0.69)
prob_of_nogun_and_college
```

Answer: The probability will be 0.3174

## Question 3

*Points: 5*

I roll a fair six-sided dice and I get a number greater than three. What is the probability that the next two dice rolls will also generate numbers greater than three?

Answer: Probability is 25%

## Question 4

*Points: 10*

Assume that in the United States, 15% of people are smokers and, of smokers, 60% smoke at least one pack a day. Among those who smoke at least a pack a day, 60% will develop lung cancer at some point in their lives while, in the rest of the p opulation (which includes smokers who smoke less than a pack a day and non-smokers), only 10% will. What is the probability that a person who develops lung cancer was a smoker who smoked at least a pack a day?

Answer: Probability is 0.3724

## Question 5

*Points: 5*

Assume that in the world, 60% of countries are democracies and 80% of countries have ratified the Rome Statute for the International Criminal Court. Assume further that 90% of democracies have ratified the Rome Accords. What is the probability that a country is either a democracy or has ratified the Rome Accords?

```{r}

p_D <- 0.60         
p_R <- 0.80          
p_R_given_D <- 0.90  

p_D_and_R <- p_D * p_R_given_D

p_D_or_R <- p_D + p_R - p_D_and_R

print(p_D_or_R)


```

Answer: Probability is 0.86
